Alone Together: Compositional Reasoning and Inference for Weak Isolation

### 2017-10-21 implementation.tex
Edits.

We have implemented our DSL to define transactions as monadic
computations in OCaml (modulo some syntactic sugar), and our automatic
reasoning framework as an extra frontend pass (called \tool) in the ocamlc
4.03 compiler\footnote{The source code is available at available at
\emph{https://goo.gl/nmUXZK}}. The input to \tool is a program in our DSL that
describes the schema of the database as a collection of OCaml type
definitions, and transactions as OCaml functions, whose top-level
expression is an application of the \C{atomically\_do} combinator to a
function of type $a \rightarrow \C{DB}\;b$, for some (base) types $a$
and $b$. For instance, TPC-C's schema from Fig.~\ref{fig:schema} can
be described via the OCaml type definitions shown in
Fig.~\ref{fig:ocaml-schema}.  \tool also requires a specification of
the program in the form of a collection of guarantees ($G$), one per
transaction, and an invariant $I$ that is a conjunction of the
integrity constraints on the database. An auxiliary DSL that includes
the first-order logic (FOL) combinators has been implemented for this
purpose. \tool's verification pass follows OCaml's type checking pass,
hence the concrete artifact of verification is OCaml's typed AST. The
tool is already equipped with  an axiomatization of PostgreSQL and
MySQL's isolation levels expressed in our FOL DSL. Other data stores
can be similarly axiomatized. The concrete result of verification is
an assignment of an isolation level of the selected data store to each
transaction in the program.

At the top-level, \tool runs a loop that picks an unverified
transaction and progressively strengthens its isolation level until it
passes verification. If the selected data store provides a
serializable isolation level, and if the program is sequentially
correct, then the verification is guaranteed to succeed. Within the
loop, the \tool first computes the various rely relations needed for
verification ($R$, $\R_l$, and $\R_c$). It then traverses the AST of a
transaction, applying the inference rules to construct a state
transformer, check its stability, and weaken it ($\stabilize{\cdot}$)
if it is not stable. The result of traversing the transaction's AST is
therefore a state transformer ($\F$) that is stable w.r.t $R_l$, which
is also stabilized against $\R_c$ (using $\stabilize{\cdot}$), and
then checked against the transaction's stated guarantee ($G$). If the
check passes, then the guarantee is verified to check if it preserves
the invariant $I$. The successful result from both the checks results
in the transaction being certified correct under the current choice of
its isolation level. Successful verification of all transactions
concludes the top-level execution, returning the inferred isolation
levels as its output.

>>>>

We have implemented our DSL to define transactions as monadic
computations in OCaml (modulo some syntactic sugar), and our automatic
reasoning framework as an extra frontend pass (called \tool) in the
ocamlc 4.03 compiler\footnote{The source code is available at
available at \emph{https://goo.gl/nmUXZK}}. The input to \tool is a
program in our DSL that describes the schema of the database as a
collection of OCaml type definitions, and transactions as OCaml
functions, whose top-level expression is an application of the
\C{atomically\_do} combinator. For instance, TPC-C's schema from
Fig.~\ref{fig:schema} can be described via the OCaml type definitions
shown in Fig.~\ref{fig:ocaml-schema}.  \tool also requires a
specification of the program in the form of a collection of guarantees
($G$), one per transaction, and an invariant $I$ that is a conjunction
of the integrity constraints on the database. An auxiliary DSL that
includes the first-order logic (FOL) combinators has been implemented
for this purpose. \tool's verification pass follows OCaml's type
checking pass, hence the concrete artifact of verification is OCaml's
typed AST. The tool is already equipped with  an axiomatization of
PostgreSQL and MySQL's isolation levels expressed in our FOL DSL.
Other data stores can be similarly axiomatized. The concrete result of
verification is an assignment of an isolation level of the selected
data store to each transaction in the program.

At the top-level, \tool runs a loop that picks an unverified
transaction and progressively strengthens its isolation level until it
passes verification. If the selected data store provides a
serializable isolation level, and if the program is sequentially
correct, then the verification is guaranteed to succeed. Within the
loop, the \tool first computes the various rely relations needed for
verification ($R$, $\R_l$, and $\R_c$). It then traverses the AST of a
transaction, applying the inference rules to construct a state
transformer, check its stability, and weaken it ($\stabilize{\cdot}$)
if it is not stable. The result of traversing the transaction's AST is
therefore a state transformer ($\F$) that is stable w.r.t $\R_l$, which
is also stabilized against $\R_c$ (using $\stabilize{\cdot}$), and
then checked against the transaction's stated guarantee ($G$). If the
check passes, then the guarantee is verified to check if it preserves
the invariant $I$. The successful result from both the checks results
in the transaction being certified correct under the current choice of
its isolation level. Successful verification of all transactions
concludes the top-level execution, returning the inferred isolation
levels as its output.

### 2017-10-21 inference.tex
Edits.

literatl expressions. For e.g., the set $\{1,2\}$ can be written as

@@

  \stable(\R,\F) & \Leftrightarrow & \forall \stl,\stg,\stg',\bar{v}.~

@@

where $\bar{v}$ are the variables that occur free in $\F$; this is

@@

The proofs for \C{SELECT}, sequencing, and conditionals use use the
inductive hypothesis to infer the RG-judgements present in the
premises of their corresponding RG-rules. The interesting case is the
\C{FOREACH} statement, for which we use the loop invariant
$\psi(\stl, \stg) \Leftrightarrow \stl = s \cup (y \bind (\lambda
z.~\F(s, \stg)))$, (where assuming that $c$ is the body of the loop,
$c \elabsto \F$) to prove all the premises of
\rulelabel{RG-Foreach}. Using the same notation as the rule
$\rulelabel{RG-Foreach}$, $y$ refers to the records already processed
in previous iterations of the loop, while $z$ refers to the record
being processed in the current iteration.  At the beginning of the
loop $[\phi/y]\psi(\stl, \stg)$ just reduces to $\stl = s$ which is
implied by the pre-condition $P$. From the inductive hypothesis, we
can infer that each iteration corresponds to the application of
$F$. Since all iterations are assumed to be independent of each other,
and $z$ is bound to a record in $x$ for each iteration, we conclude
that at the end of every iteration, the loop invariant $[y \cup
  \{z\}/y]\psi$ will be satisfied.

@@

The algorithm ($\mssemof{\cdot}{\cdot}$) shown in Fig.~\ref{fig:logic}
translates an $\SL$ expression ($s$) to FOL. The translation is based
on encoding a set of element type $T$ as a unary predicate on $T$.
The predicate is represented as a meta function that accepts an
$x:T$ and returns a quantifier-free proposition that evaluates to
true ($\top$) if and only if $x$ is present in the set. Alternatively,
translation may also encode the set as a function in the logic itself,
in which case a quantified proposition constraining the function is
also generated. For instance, consider the set $\{1,2\}$. The
predicate describing the set can be encoded as the function $\lambda
\v. \v=1 \vee \v=2$, with no further constraints, or it can be encoded as the function $\lambda \v. g(\v)$
with an associated constraint ($\phi\in\Prop^1$) defining the
uninterpreted boolean functon $g$:
\begin{smathpar}
  \begin{array}{lcl}
    \phi & = & \forall \nu.~g(\nu) \Leftrightarrow \nu=1 \vee \nu=2
  \end{array}
\end{smathpar}
The translation adopts one or the other approach, depending on the
need. For uniformity, we consider the encoding of a set as pair
($\phi$,$\G$), where $\G$ is a meta function, and $\phi$ is a FOL
formula constraining any uninterpreted functions used in $G$.

@@

The translation algorithm (Fig.~\ref{fig:logic}) takes free variables
into account. Given a set expression $s\in\SL$, whose (possible) free
variables are $\nubar$ in the order of their introduction (top-most
binder first), $\mssemof{\nubar}{s}$ returns the encoding of $s$ as
$(\phi,\G)$.  The meta function $\G$ is a predicate indexed by the
(possible) free variables of $s$, thus, its arity is $|\nubar| + 1$.
Note that $\nubar$ is only a sequence of variables introduced by the
enclosing binders of $s$, and not all may actually occur free in $s$.
Nonetheless, its predicate $\G$ is always indexed by $|\nubar|$ free
variables for uniformity. The translation encodes database state as an
uninterpreted sort. Considering that the state is actually a set of
records, we define an uninterpreted relation $\in$ to relate records
and states. Thus, a variable set expression $\stg$ denoting a database
state is encoded as the predicate $\lambda (\vbar,r).~r\in\stg$, where
$|\vbar|=|\nubar|$ (predicates are uncurried for simplicity; $\vbar$
is a comma-separated sequence; $r\not\in\SL$ is a special variable).
The constraints associated with the encoding of a state are trivial
(denoted $\top$). The set literal expression $\{x\,|\, \varphi\}$ is encoded
straightforwardly. The conditional set expression is encoded as an
if-then-else predicate in FOL, where the predicates on true and false
branches are computed from the set subexpressions $s_1$ and $s_2$,
respectively. The conjunction of constraints $\phi_1$ and $\phi_2$,
from $\mssemof{\nubar}{s_1}$ and $\mssemof{\nubar}{s_2}$ (resp.), is
propagated upwards as the constraint of the conditional expression.
Set union expression is encoded similarly.

>>>>

literal expressions. For e.g., the set $\{1,2\}$ can be written as

@@

  \stable(\R,\F) & \Leftrightarrow & \forall \stl,\stg,\stg',\nubar.~

@@

where $\nubar$ are the variables that occur free in $\F$; this is

@@

The proofs for \C{SELECT}, sequencing, and conditionals use use the
inductive hypothesis to infer the RG-judgments present in the
premises of their corresponding RG-rules. The interesting case is the
\C{FOREACH} statement, for which we use the loop invariant
$\psi(\stl, \stg) \Leftrightarrow \stl = s \cup (y \bind (\lambda
z.~\F(s, \stg)))$, (where assuming that $c$ is the body of the loop,
$c \elabsto \F$) to prove all the premises of
\rulelabel{RG-Foreach}. Using the same notation as the rule
$\rulelabel{RG-Foreach}$, $y$ refers to the records already processed
in previous iterations of the loop, while $z$ refers to the record
being processed in the current iteration.  At the beginning of the
loop $[\phi/y]\psi(\stl, \stg)$ just reduces to $\stl = s$ which is
implied by the pre-condition $P$. From the inductive hypothesis, we
can infer that each iteration corresponds to the application of
$F$. Since all iterations are assumed to be independent of each other,
and $z$ is bound to a record in $x$ for each iteration, we conclude
that at the end of every iteration, the loop invariant $[y \cup
  \{z\}/y]\psi$ will be satisfied.

@@

The algorithm ($\mssemof{\cdot}{\cdot}$) shown in Fig.~\ref{fig:logic}
translates an $\SL$ expression ($s$) to FOL. The translation is based
on encoding a set of element type $T$ as a unary predicate on $T$.
The predicate is represented as a meta function that accepts an $x:T$
and returns a quantifier-free proposition that evaluates to true
($\top$) if and only if $x$ is present in the set. Alternatively,
translation may also encode the set as a predicate in the logic
itself, in which case a quantified proposition constraining the
predicate is also generated. For instance, consider the set $\{1,2\}$.
The predicate describing the set can be encoded as the function
$\lambda \v. \v=1 \vee \v=2$, with no further constraints, or it can
be encoded as the function $\lambda \v. g(\v)$ with an associated
constraint ($\phi\in\Prop^1$) defining the uninterpreted boolean
function (i.e., uninterpreted predicate) $g$:
\begin{smathpar}
  \begin{array}{lcl}
    \phi & = & \forall \nu.~g(\nu) \Leftrightarrow \nu=1 \vee \nu=2
  \end{array}
\end{smathpar}
The translation adopts one or the other approach, depending on the
need. For uniformity, we consider the encoding of a set as pair
($\phi$,$\G$), where $\G$ is a meta function, and $\phi$ is a FOL
formula constraining any uninterpreted predicates used in $G$.

@@

The translation algorithm (Fig.~\ref{fig:logic}) takes free variables
into account. Given a set expression $s\in\SL$, whose (possible) free
variables are $\nubar$ in the order of their introduction (top-most
binder first), $\mssemof{\nubar}{s}$ returns the encoding of $s$ as
$(\phi,\G)$.  The meta function $\G$ is a predicate indexed by the
(possible) free variables of $s$, thus, its arity is $|\nubar| + 1$.
Note that $\nubar$ is only a sequence of variables introduced by the
enclosing binders of $s$, and not all may actually occur free in $s$.
Nonetheless, its predicate $\G$ is always indexed by $|\nubar|$ free
variables for uniformity. The translation encodes database state as an
uninterpreted sort. Considering that the state is actually a set of
records, we define an uninterpreted relation ``$\in$'' to relate records
and states. Thus, a variable set expression $\stg$ denoting a database
state is encoded as the predicate $\lambda (\vbar,r).~r\in\stg$, where
$|\vbar|=|\nubar|$ (predicates are uncurried for simplicity; $\vbar$
is a comma-separated sequence; $r\not\in\SL$ is a special variable).
The constraints associated with the encoding of a state are trivial
(denoted $\top$). The set literal expression $\{x\,|\, \varphi\}$ is encoded
straightforwardly. The conditional set expression is encoded as an
if-then-else predicate in FOL, where the predicates on true and false
branches are computed from the set subexpressions $s_1$ and $s_2$,
respectively. The conjunction of constraints $\phi_1$ and $\phi_2$,
from $\mssemof{\nubar}{s_1}$ and $\mssemof{\nubar}{s_2}$ (resp.), is
propagated upwards as the constraint of the conditional expression.
Set union expression is encoded similarly.

### 2017-10-21 opsem.tex
Edits.

Fig.~\ref{fig:txnimp} shows the syntax and small-step semantics of
\txnimp, a core language that we use to formalize our intuitions about
reasoning under weak isolation. Variables ($x$), integer and Boolean
constants ($k$), records ($r$) of named constants, sets ($s$) of such
records, arithmetic and boolean expressions ($e_1 \odot e_2$), and
record expressions ($[\bar{f}=\bar{e}]$) constitute the syntactic
class of expressions ($e$). Commands ($c$) include $\cskip$,
conditional statements, \C{LET} constructs to bind names, \C{FOREACH}
loops, SQL statements, their sequential composition ($c_1;c_2$),
transactions ($\ctxn{i}{\I}{c}$) and their parallel composition
($c_1\,||\,c_2$). Each transaction is assumed to have a unique
identifier $i$, and executes at the top-level; our semantics does not
support nested transactions. The $\I$ in the \C{TXN} block syntax is
the transaction's isolation specification, whose purpose is explained
below.  Certain terms that only appear at run-time are also present in
$c$.  These include a {\sf txn} block tagged with sets ($\stl$ and
$\stg$) of records representing local and global database state, and a
runtime {\sf foreach} expression that keeps track of the set ($s_1$)
of items already iterated, and the set ($s_2$) of items yet to be
iterated. Note that the surface-level syntax of the \C{FOREACH}
command shown here is slightly different from the one used in previous
sections; its higher-order function has two arguments, $y$ and $z$,
which are invoked (during the reduction) with the set of
already-iterated items, and the current item, respectively. This form
of \C{FOREACH} lends itself to inductive reasoning that will be useful
for verification (Sec.~\ref{sec:reasoning}).  Our language ensures
that all effectful actions are encapsulated within database commands,
and that all shared state among processes are only manipulated via
transactions and its supported operations.  In particular, we do not
consider programs in which objects resident on e.g., the OCaml heap
are concurrently manipulated by OCaml expressions as well as database
actions.

>>>>

Fig.~\ref{fig:txnimp} shows the syntax and small-step semantics of
\txnimp, a core language that we use to formalize our intuitions about
reasoning under weak isolation. Variables ($x$), integer and Boolean
constants ($k$), records ($r$) of named constants, sets ($s$) of such
records, arithmetic and boolean expressions ($e_1 \odot e_2$), and
record expressions ($\langle \bar{f}=\bar{e} \rangle$) constitute the
syntactic class of expressions ($e$). Commands ($c$) include $\cskip$,
conditional statements, \C{LET} constructs to bind names, \C{FOREACH}
loops, SQL statements, their sequential composition ($c_1;c_2$),
transactions ($\ctxn{i}{\I}{c}$) and their parallel composition
($c_1\,||\,c_2$). Each transaction is assumed to have a unique
identifier $i$, and executes at the top-level; our semantics does not
support nested transactions. The $\I$ in the \C{TXN} block syntax is
the transaction's isolation specification, whose purpose is explained
below.  Certain terms that only appear at run-time are also present in
$c$.  These include a {\sf txn} block tagged with sets ($\stl$ and
$\stg$) of records representing local and global database state, and a
runtime {\sf foreach} expression that keeps track of the set ($s_1$)
of items already iterated, and the set ($s_2$) of items yet to be
iterated. Note that the surface-level syntax of the \C{FOREACH}
command shown here is slightly different from the one used in previous
sections; its higher-order function has two arguments, $y$ and $z$,
which are invoked (during the reduction) with the set of
already-iterated items, and the current item, respectively. This form
of \C{FOREACH} lends itself to inductive reasoning that will be useful
for verification (Sec.~\ref{sec:reasoning}).  Our language ensures
that all effectful actions are encapsulated within database commands,
and that all shared state among processes are only manipulated via
transactions and its supported operations.  In particular, we do not
consider programs in which objects resident on e.g., the OCaml heap
are concurrently manipulated by OCaml expressions as well as database
actions.

### 2017-10-21 reasoning.tex
Edits.

We now describe a proof system that lets us prove the correctness of a
\txnimp program $c$ w.r.t its high-level consistency conditions $I$,
on an implementation that satisfies the isolation specifications
($\I$) of its transactions\footnote{Note the difference between $I$
  and $\I$. The former constitute \emph{proof} \emph{obligations} for
  the programmer, whereas the latter describes a transaction's
  \emph{assumptions} about the operational characteristics of the
  underlying system.}.  Our proof system is essentially an adaptation
of a rely-guarantee reasoning framework~\cite{rgjones} to the setting
of weakly isolated database transactions.  The primary challenge in
the formulation deals with how we relate a transaction's isolation
specification ($\I$) to its rely relation ($R$) that describes the
transaction's environment, so that interference is considered only
insofar as allowed by the isolation level.  Another characteristic of
the transaction setting that affects the structure of the proof system
is atomicity; we do not permit a transaction's writes to be visible
until it commits.  In the context of rely-guarantee, this means that
the transaction's guarantee ($G$) should capture the aggregate effect
of a transaction, and not its individual writes.  While shared memory
\C{atomic} blocks also have the same characteristic, the fact that
transactions are weakly-isolated introduces surprising complexity.
Unlike an \C{atomic} block, the effect of a transaction is \emph{not}
a sequential composition of the effects of its statements because each
statement can witness a potentially different version of the state.

@@

To obtain an equivalent form in our setting, we require an additional
condition on the isolation specification, which we call the
\emph{stability condition on $\I$}.  The condition requires $\I$ to
allow interferences $R^{*}(\stg,\stg'')$, for two database states
$\stg$ and $\stg''$, only if it also allows interference for every
prefix of $R^{*}(\stg,\stg'')$. In other words, if $\I$ disallows
interference from $\stg$ to $\stg'$, then an $R$-step from $\stg'$ to
$\stg''$ should not make the interference from $\stg$ to $\stg''$
valid.  This stability condition on $\I$ is defined formally thus:
\begin{smathpar}
\begin{array}{lcl}
  \stable(R,\I) & \Leftrightarrow & \forall \stl,\stg,\stg',\stg''.~
  \neg\I(\stl,\stg,\stg') \conj R(\stg',\stg'') \Rightarrow
  \neg\I(\stl,\stg,\stg'')
\end{array}
\end{smathpar}
It can be easily verified that the above stability condition is
satisfied by the isolation axioms from Sec.~\ref{sec:isolation}. For
instance, $\I_{ss}$, the snapshot axiom, is stable because if
$\I_{ss}$ is invalid ($\neg\I_{ss}$), then an interference has already
modified a record, and no further interference will restore the
original record, because the original record bears the id of a
transaction that has already committed. We can now generalize the
stability condition for any arbitrary assertion $\phi$ by only
considering those environment steps which in addition to preserving
$\phi$, also preserve the condition that the transaction can continue
to take its next step, i.e. the condition $\I$ holds on both the
beginning and the end states.
 \begin{smathpar}
   \forall \stl, \stg, \stg', \stg_1. \phi(\stl, \stg) \wedge \I(\stl, \stg_1, \stg) \wedge R^{*}(\stg, \stg') \wedge \I(\stl, \stg_1, \stg')
   \Rightarrow \phi(\stl, \stg')
\end{smathpar}
Here, we use a separate notation ($\stg_1$) for the global database
state that the transaction witnessed during its last step. $\phi$ does
not constrain this global database state, but the stability of $\I$
ensures that any database state constrained by $\phi$ must also obey
$\I$. We can now replace the transitive closure in the rule with the
single step relation $R$, backed by the inductive argument for
stability. Pushing the universal quantifier for $\stg_1$ inside the
antecedent and separating the predicates involving $\stg_1$, we can
rewrite the condition in the form below, by defining a new ternary
relation $R \backslash \I$, which can be thought of as $R$ modulo
$\I$:
\begin{smathpar}
\begin{array}{lcl}
\stable((R \backslash \I), \phi) & \Leftrightarrow & \forall \stl,\stg,\stg'. \phi(\stl,\stg) \wedge (R \backslash \I)(\stl,\stg,\stg') \Rightarrow \phi(\stl,\stg')\\
(R \backslash \I)(\stl, \stg, \stg') & \Leftrightarrow & \exists \stg_1. R(\stg, \stg') \wedge \I(\stl, \stg_1, \stg) \wedge \I(\stl, \stg_1, \stg')
\end{array}
\end{smathpar}
This condition often significantly simplifies the form of $R
\backslash \I$ irrespective of $R$. For example, when a transaction is
executed against a snapshot of the database (i.e. $\mathbb{I}_{ss}$),
$R \backslash \I_{ss}$ will be the identity function, since if $\stg
\neq \stg'$, there does not exist any $\stg_1$ which satisfies the
constraint for $(R \backslash \I_{ss})(\stl,\stg,\stg')$.

@@

\begin{definition}[\bfseries Semantics of RG judgments]
\label{def:rg-semantics}
The semantics of the global and transaction-local RG judgments are
defined thus:
\begin{smathpar}
\begin{array}{lclr}
\R \vdash \hoare{P}{c}{Q} & \defeq & \forall \stl,\stl',\stg,\stg'.~ P(\stl,\stg) \conj (\tbox{c}_i,\stl,\stg) \rstepssto{n}
  (\tbox{\cskip}_i, \stl',\stg')
  \Rightarrow Q(\stl',\stg') \\
\rg{I,R}{c}{G,I} & \defeq &  \forall \stg.\, I(\stg)
  \Rightarrow (\forall \stg'.\; (c,\stg) 
    \rstepssto{n} (\cskip,\stg') \Rightarrow I(\stg')) \\
&&\hspace*{0.5in}\conj \texttt{step-guaranteed}(R,G,c,\stg)\\
\end{array}
\end{smathpar}

>>>>

We now describe a proof system that lets us prove the correctness of a
\txnimp program $c$ w.r.t its high-level consistency conditions $I$,
on an implementation that satisfies the isolation specifications
($\I$) of its transactions\footnote{Note the difference between $I$
  and $\I$. The former constitute \emph{proof} \emph{obligations} for
  the programmer, whereas the latter describes a transaction's
  \emph{assumptions} about the operational characteristics of the
  underlying system.}.  Our proof system is essentially an adaptation
of a rely-guarantee reasoning framework~\cite{rgjones} to the setting
of weakly isolated database transactions.  The primary challenge in
the formulation deals with how we relate a transaction's isolation
specification ($\I$) to its rely relation ($R$) that describes the
transaction's environment, so that interference is considered only
insofar as allowed by the isolation level.  Another characteristic of
the transaction setting that affects the structure of the proof system
is atomicity; we do not permit a transaction's writes to be visible
until it commits.  In the context of rely-guarantee, this means that
the transaction's guarantee ($G$) should capture the aggregate effect
of a transaction, and not its individual writes.  While shared memory
\C{atomic} blocks also have the same characteristic, the fact that
transactions are weakly-isolated introduces non-trivial complexity.
Unlike an \C{atomic} block, the effect of a transaction is \emph{not}
a sequential composition of the effects of its statements because each
statement can witness a potentially different version of the state.

@@

To obtain an equivalent form in our setting, we require an additional
condition on the isolation specification, which we call the
\emph{stability condition on $\I$}.  The condition requires $\I$ to
allow interferences $R^{*}(\stg,\stg'')$, for two database states
$\stg$ and $\stg''$, only if it also allows interference for every
prefix of $R^{*}(\stg,\stg'')$. In other words, if $\I$ disallows
interference from $\stg$ to $\stg'$, then an $R$-step from $\stg'$ to
$\stg''$ should not make the interference from $\stg$ to $\stg''$
valid.  This stability condition on $\I$ is defined formally thus:
\begin{smathpar}
\begin{array}{lcl}
  \stable(R,\I) & \Leftrightarrow & \forall \stl,\stg,\stg',\stg''.~
  \neg\I(\stl,\stg,\stg') \conj R(\stg',\stg'') \Rightarrow
  \neg\I(\stl,\stg,\stg'')
\end{array}
\end{smathpar}
It can be easily verified that the above stability condition is
satisfied by the isolation axioms from Sec.~\ref{sec:isolation}. For
instance, $\I_{ss}$, the snapshot axiom, is stable because if
$\I_{ss}$ is invalid ($\neg\I_{ss}$), then an interference has already
modified a record, and no further interference will restore the
original record, because new commits bear new transaction ids
different from the transaction id (\C{txn} field) of the original
record. We can now generalize the stability condition for any
arbitrary assertion $\phi$ by only considering those environment steps
which in addition to preserving $\phi$, also preserve the condition
that the transaction can continue to take its next step, i.e. the
condition $\I$ holds on both the beginning and the end states.
 \begin{smathpar}
   \forall \stl, \stg, \stg', \stg_1. \phi(\stl, \stg) \wedge \I(\stl, \stg_1, \stg) \wedge R^{*}(\stg, \stg') \wedge \I(\stl, \stg_1, \stg')
   \Rightarrow \phi(\stl, \stg')
\end{smathpar}
Here, we use a separate notation ($\stg_1$) for the global database
state that the transaction witnessed during its last step. $\phi$ does
not constrain this global database state, but the stability of $\I$
ensures that any database state constrained by $\phi$ must also obey
$\I$. We can now replace the transitive closure in the rule with the
single step relation $R$, backed by the inductive argument for
stability. Pushing the universal quantifier for $\stg_1$ inside the
antecedent and separating the predicates involving $\stg_1$, we can
rewrite the condition in the form below, by defining a new ternary
relation $R \backslash \I$, which can be thought of as $R$ modulo
$\I$:
\begin{smathpar}
\begin{array}{lcl}
\stable((R \backslash \I), \phi) & \Leftrightarrow & \forall \stl,\stg,\stg'. \phi(\stl,\stg) \wedge (R \backslash \I)(\stl,\stg,\stg') \Rightarrow \phi(\stl,\stg')\\
(R \backslash \I)(\stl, \stg, \stg') & \Leftrightarrow & \exists \stg_1. R(\stg, \stg') \wedge \I(\stl, \stg_1, \stg) \wedge \I(\stl, \stg_1, \stg')
\end{array}
\end{smathpar}
This condition often significantly simplifies the form of $R
\backslash \I$ irrespective of $R$. For example, when a transaction is
executed against a snapshot of the database (i.e. $\mathbb{I}_{ss}$),
$R \backslash \I_{ss}$ will be the identity function, since if $\stg
\neq \stg'$, there does not exist any $\stg_1$ which satisfies the
constraint for $(R \backslash \I_{ss})(\stl,\stg,\stg')$.

@@

\begin{definition}[\bfseries Semantics of RG judgments]
\label{def:rg-semantics}
The semantics of the global and transaction-local RG judgments are
defined thus:
\begin{smathpar}
\begin{array}{lclr}
\R \vdash \hoare{P}{c}{Q} & \defeq & \forall
  \stl,\stl',\stg,\stg',\nubar.~ P(\stl,\stg) \conj (\tbox{c}_i,\stl,\stg) \rstepssto{n}
  (\tbox{\cskip}_i, \stl',\stg')
  \Rightarrow Q(\stl',\stg') & \nubar = \texttt{freevars}(c)\\
\rg{I,R}{c}{G,I} & \defeq &  \forall \stg.\, I(\stg)
  \Rightarrow (\forall \stg'.\; (c,\stg) 
    \rstepssto{n} (\cskip,\stg') \Rightarrow I(\stg')) \\
&&\hspace*{0.6in}\conj \texttt{step-guaranteed}(R,G,c,\stg)\\
\end{array}
\end{smathpar}
